### 2025-02-13 paper/paper_copy.tex
Initial commit

\subsection*{Single Agent Implementation}

To validate our approach, we first implement a single-agent version of the learning framework. This simplified setting allows us to verify the learning dynamics and compare them against the theoretical bounds established in Brandl (2024). The implementation follows the binary state case ($|\Omega| = 2$) with symmetric signals of accuracy $q > 0.5$.

\subsubsection*{Metrics Tracking}
We track two key metrics throughout the learning process:

\begin{itemize}
    \item \textit{Mistake Rate}: The empirical probability of the agent choosing an incorrect action, computed as the running average of mistakes:
    \[
        P(a_t \neq a_\omega) = \frac{1}{t}\sum_{s=1}^t \mathbf{1}_{\{a_s \neq a_\omega\}}
    \]

    \item \textit{Learning Rate}: Following the paper's definition, we compute:
    \[
        r = \liminf_{t \to \infty} -\frac{1}{t} \log P(a_t \neq a_\omega)
    \]
    To approximate this limit, we compute the learning rate over exponentially increasing time windows and take the minimum rate, ensuring we capture the asymptotic behavior.
\end{itemize}

\subsubsection*{Environment}
The environment implements the binary state case with the following components:

\begin{itemize}
    \item A fixed true state $\omega \in \{0,1\}$ drawn uniformly at random at initialization
    \item Signal generation with accuracy $q$:
    \[
        P(s_t = \omega) = q, \quad P(s_t \neq \omega) = 1-q
    \]
    \item The reward function derived in the binary case section:
    \[
        v(a,s) = \frac{q \cdot \mathbf{1}_{\{a = s\}} - (1-q) \cdot \mathbf{1}_{\{a \neq s\}}}{2q-1}
    \]
\end{itemize}

\subsubsection*{Neural Network Architecture}
The agent employs an actor-critic architecture with both networks using GRU cells to process the history of signals. The GRU implementation follows the equations presented in the RNN Architecture section, with the following specifications:

\paragraph{Actor Network}
The actor network maps the signal history to a policy over actions:
\begin{align*}
    h_t &= \text{GRU}(s_t, h_{t-1}) \\
    \pi(a_t|h_t) &= \text{softmax}(W_\pi h_t + b_\pi)
\end{align*}

\paragraph{Critic Network}
The critic network estimates state values using the same GRU architecture:
\begin{align*}
    h_t &= \text{GRU}(s_t, h_{t-1}) \\
    V(h_t) &= W_v h_t + b_v
\end{align*}

\subsubsection*{Training Process}
The agent is trained using a variant of the Advantage Actor-Critic (A2C) algorithm. The complete training procedure is detailed in Algorithm \ref{alg:single-agent-training}.

The algorithm implements the A2C training loop, processing signals through GRU-based actor and critic networks while tracking learning metrics. The actor network learns a policy that minimizes mistakes in predicting the true state, while the critic network estimates state values to compute advantages for policy updates.

\subsubsection*{Theoretical Bounds}
For the binary case with symmetric signals, the theoretical bound on the learning rate is:
\[
    r_{bdd} = -\log(1-q_{bdd}) = -\log(1-(2q-1))
\]
where $q_{bdd} = 2q-1$ is the bound on the probability of choosing the correct action. This provides a benchmark against which we can evaluate the empirical learning rate of our implementation.

\subsubsection*{Empirical Results}
The implementation is evaluated over multiple training runs with different random seeds. For each run, we track:
\begin{itemize}
    \item The evolution of the mistake rate over time
    \item The empirical learning rate compared to the theoretical bound
    \item The convergence of the policy to the optimal action
\end{itemize}

>>>>

\pagebreak

\section*{Appendix A: Theoretical Background}

\subsection*{Deep Reinforcement Learning Framework}

The foundation of our deep reinforcement learning approach builds upon the policy gradient theorem \citep{sutton2000policy} and its extensions to deep neural networks \citep{mnih2015human}. In the context of multi-agent systems, we specifically utilize the actor-critic framework \citep{konda2000actor}, which has been shown to be particularly effective in handling the increased complexity of multi-agent environments \citep{lowe2017multi}.

\subsection*{Policy Gradient Methods}

The core objective in reinforcement learning is to maximize the expected cumulative discounted reward,
\[
J(\phi) = \mathbb{E}_{\tau \sim \pi_\phi}[\sum_{t=0}^{\infty} \gamma^t r_t]
\]
where $\tau$ represents a trajectory, and $\pi_\phi$ is the policy parameterized by $\phi$. The policy gradient theorem \citep{sutton2000policy} provides the gradient of this objective,
\[
\nabla_\phi J(\phi) = \mathbb{E}_{\tau \sim \pi_\phi}[\sum_{t=0}^{\infty} \nabla_\phi \log \pi_\phi(a_t|s_t) Q^{\pi_\phi}(s_t, a_t)]
\]
In practice, we use the advantage function $A(s_t, a_t)$ instead of the Q-function to reduce variance \citep{schulman2015high}:

\[
\nabla_\phi J(\phi) = \mathbb{E}_{\tau \sim \pi_\phi}[\sum_{t=0}^{\infty} \nabla_\phi \log \pi_\phi(a_t|s_t) A(s_t, a_t)]
\]

\subsection*{Value Function Approximation}
The value function approximation follows the temporal difference (TD) learning framework \citep{sutton2018reinforcement}. The value network parameters $\phi$ are updated to minimize the mean squared TD error:

\[
\mathcal{L}(\phi) = \mathbb{E}_{(s_t, a_t, r_t, s_{t+1}) \sim \mathcal{D}}[(r_t + \gamma V_\phi(s_{t+1}) - V_\phi(s_t))^2]
\]

where $\mathcal{D}$ represents the replay buffer of experiences. This approach has been shown to be particularly effective when combined with deep neural networks \citep{mnih2015human}.

\subsubsection*{LSTM Implementation}

In practice, we use Long Short-Term Memory (LSTM) cells to better capture long-term dependencies in the information sets. The LSTM update equations for each agent $i$ are:
\begin{align}
    f^i_t &= \sigma(W^i_f [h^i_{t-1}, s^i_t, \mathbf{a}_t] + b^i_f) \\
    \iota^i_t &= \sigma(W^i_\iota [h^i_{t-1}, s^i_t, \mathbf{a}_t] + b^i_\iota) \\
    \tilde{c}^i_t &= \tanh(W^i_c [h^i_{t-1}, s^i_t, \mathbf{a}_t] + b^i_c) \\
    c^i_t &= f^i_t \odot c^i_{t-1} + \iota^i_t \odot \tilde{c}^i_t \\
    o^i_t &= \sigma(W^i_o [h^i_{t-1}, s^i_t, \mathbf{a}_t] + b^i_o) \\
    h^i_t &= o^i_t \odot \tanh(c^i_t)
\end{align}
where $c^i_t$ is the cell state that allows for longer-term memory preservation.

\subsection*{Attention-Enhanced RNN Architecture}

We extend the base RNN architecture by incorporating both self-attention and cross-attention mechanisms to better capture temporal dependencies and agent interactions. The attention mechanisms allow each agent to selectively focus on relevant historical information and other agents' actions.

\subsubsection*{Self-Attention Mechanism}
For each agent $i$, we implement a self-attention mechanism that operates over a window of $K$ previous hidden states. Let $H_t^i = [h_{t-K+1}^i, ..., h_t^i] \in \mathbb{R}^{d \times K}$ be the matrix of $K$ previous hidden states. The self-attention operation is defined as:

\begin{align*}
Q_t^i &= W_Q^i h_t^i + b_Q^i & \text{(query)} \\
K_t^i &= W_K^i H_t^i + b_K^i & \text{(keys)} \\
V_t^i &= W_V^i H_t^i + b_V^i & \text{(values)} \\
\alpha_t^i &= \text{softmax}\left(\frac{(Q_t^i)^\top K_t^i}{\sqrt{d}}\right) & \text{(attention weights)} \\
c_t^i &= V_t^i \alpha_t^i & \text{(context vector)}
\end{align*}

where $W_Q^i, W_K^i, W_V^i$ are learned weight matrices and $b_Q^i, b_K^i, b_V^i$ are bias vectors. The scaling factor $\sqrt{d}$ prevents the dot products from growing too large in magnitude.

\subsubsection*{Cross-Attention for Agent Interactions}
To better capture agent interactions, we implement a cross-attention mechanism that allows each agent to attend to other agents' hidden states. For agent $i$, let $H_t^{-i} = [h_t^1, ..., h_t^{i-1}, h_t^{i+1}, ..., h_t^n] \in \mathbb{R}^{d \times (n-1)}$ be the matrix of other agents' current hidden states. The cross-attention operation is:

\begin{align*}
Q_t^{i,c} &= W_{Q,c}^i h_t^i + b_{Q,c}^i & \text{(cross query)} \\
K_t^{i,c} &= W_{K,c}^i H_t^{-i} + b_{K,c}^i & \text{(cross keys)} \\
V_t^{i,c} &= W_{V,c}^i H_t^{-i} + b_{V,c}^i & \text{(cross values)} \\
\beta_t^i &= \text{softmax}\left(\frac{(Q_t^{i,c})^\top K_t^{i,c}}{\sqrt{d}}\right) & \text{(cross-attention weights)} \\
m_t^i &= V_t^{i,c} \beta_t^i & \text{(cross-context vector)}
\end{align*}

\subsubsection*{Enhanced GRU Architecture}
We modify the original GRU architecture to incorporate both attention mechanisms. The updated equations become:

\begin{align*}
\hat{h}_t^i &= [h_t^i; c_t^i; m_t^i] & \text{(concatenated state)} \\
z_t^i &= \text{sigmoid}(W_z^i[\hat{h}_t^i, s_t^i, \mathbf{a}_t] + b_z^i) & \text{(update gate)} \\
\rho_t^i &= \text{sigmoid}(W_\rho^i[\hat{h}_t^i, s_t^i, \mathbf{a}_t] + b_\rho^i) & \text{(reset gate)} \\
\tilde{h}_t^i &= \tanh(W_h^i[\rho_t^i \odot \hat{h}_t^i, s_t^i, \mathbf{a}_t] + b_h^i) & \text{(candidate state)} \\
h_t^i &= (1-z_t^i) \odot \hat{h}_t^i + z_t^i \odot \tilde{h}_t^i & \text{(final update)}
\end{align*}

\subsubsection*{Multi-Head Attention Extension}
To capture different types of dependencies simultaneously, we extend both attention mechanisms to use multiple attention heads. For $L$ attention heads, each head $l \in \{1, ..., L\}$ computes its own attention weights and context vectors:

\begin{align*}
Q_{t,l}^i &= W_{Q,l}^i h_t^i + b_{Q,l}^i \\
K_{t,l}^i &= W_{K,l}^i H_t^i + b_{K,l}^i \\
V_{t,l}^i &= W_{V,l}^i H_t^i + b_{V,l}^i \\
\alpha_{t,l}^i &= \text{softmax}\left(\frac{(Q_{t,l}^i)^\top K_{t,l}^i}{\sqrt{d/L}}\right) \\
c_{t,l}^i &= V_{t,l}^i \alpha_{t,l}^i
\end{align*}

The final context vector is computed by concatenating the outputs from all heads and applying a linear transformation:

\[
c_t^i = W_O^i[c_{t,1}^i; ...; c_{t,L}^i] + b_O^i
\]

Similarly for the cross-attention mechanism:

\begin{align*}
Q_{t,l}^{i,c} &= W_{Q,c,l}^i h_t^i + b_{Q,c,l}^i \\
K_{t,l}^{i,c} &= W_{K,c,l}^i H_t^{-i} + b_{K,c,l}^i \\
V_{t,l}^{i,c} &= W_{V,c,l}^i H_t^{-i} + b_{V,c,l}^i \\
\beta_{t,l}^i &= \text{softmax}\left(\frac{(Q_{t,l}^{i,c})^\top K_{t,l}^{i,c}}{\sqrt{d/L}}\right) \\
m_{t,l}^i &= V_{t,l}^{i,c} \beta_{t,l}^i \\
m_t^i &= W_{O,c}^i[m_{t,1}^i; ...; m_{t,L}^i] + b_{O,c}^i
\end{align*}

\subsubsection*{Learning Considerations}
The addition of attention mechanisms introduces several hyperparameters that need to be tuned:
\begin{itemize}
\item History window size $K$ for self-attention
\item Number of attention heads $L$
\item Hidden dimension $d$ for attention computations
\end{itemize}

The training process remains similar to the original architecture, but with additional parameters in both actor and critic networks. The critic network now also incorporates attention mechanisms when processing the global state:

\[
Q_{\xi}(\psi_t, \mathbf{h}_t) = W_Q[\psi_t, \text{Attention}(\mathbf{h}_t), \mathbf{a}_t] + b_Q
\]

where $\text{Attention}(\mathbf{h}_t)$ applies both self and cross-attention mechanisms to the collection of all agents' hidden states.

\section*{Detailed Multi-Agent Attention PPO (MA-APPO) Algorithm}

\subsection*{Network Architectures}

\subsubsection*{1. Actor Network Architecture}

For each agent $i$, the actor network consists of:

\begin{enumerate}
\item \textbf{Input Processing Layer:}
   \[
   x_t^i = W_{\text{in}}[s_t^i; \mathbf{a}_t] + b_{\text{in}}
   \]

\item \textbf{GRU Layer with Attention:}
   \begin{align*}
   z_t^i &= \sigma(W_z^i[h_{t-1}^i, x_t^i] + b_z^i) \\
   r_t^i &= \sigma(W_r^i[h_{t-1}^i, x_t^i] + b_r^i) \\
   \tilde{h}_t^i &= \tanh(W_h^i[r_t^i \odot h_{t-1}^i, x_t^i] + b_h^i) \\
   h_t^i &= (1-z_t^i) \odot h_{t-1}^i + z_t^i \odot \tilde{h}_t^i
   \end{align*}

\item \textbf{Self-Attention Module:}
   \begin{align*}
   Q_t^i &= W_Q^i h_t^i + b_Q^i \\
   K_t^i &= W_K^i H_t^i + b_K^i \\
   V_t^i &= W_V^i H_t^i + b_V^i \\
   \alpha_t^i &= \text{softmax}\left(\frac{Q_t^i(K_t^i)^\top}{\sqrt{d_k}}\right) \\
   c_t^i &= V_t^i\alpha_t^i
   \end{align*}

\item \textbf{Cross-Attention Module:}
   \begin{align*}
   Q_t^{i,c} &= W_{Q,c}^i h_t^i + b_{Q,c}^i \\
   K_t^{i,c} &= W_{K,c}^i H_t^{-i} + b_{K,c}^i \\
   V_t^{i,c} &= W_{V,c}^i H_t^{-i} + b_{V,c}^i \\
   \beta_t^i &= \text{softmax}\left(\frac{Q_t^{i,c}(K_t^{i,c})^\top}{\sqrt{d_k}}\right) \\
   m_t^i &= V_t^{i,c}\beta_t^i
   \end{align*}

\item \textbf{Policy Head:}
   \[
   \pi_{\phi_i}(a|h_t^i) = \text{softmax}(W_\pi[h_t^i; c_t^i; m_t^i] + b_\pi)
   \]
\end{enumerate}

\subsubsection*{2. Critic Network Architecture}

The centralized critic network processes global state information:

\begin{enumerate}
\item \textbf{State Embedding:}
   \[
   e_t = W_e[\psi_t; \mathbf{h}_t] + b_e
   \]

\item \textbf{Multi-head Attention:}
   For each head $l \in \{1,\ldots,L\}$:
   \begin{align*}
   Q_{t,l} &= W_{Q,l}e_t + b_{Q,l} \\
   K_{t,l} &= W_{K,l}e_t + b_{K,l} \\
   V_{t,l} &= W_{V,l}e_t + b_{V,l} \\
   \text{head}_l &= \text{softmax}\left(\frac{Q_{t,l}K_{t,l}^\top}{\sqrt{d_k}}\right)V_{t,l}
   \end{align*}

\item \textbf{Value Head:}
   \[
   V(s_t) = W_v[\text{concat}(\text{head}_1,\ldots,\text{head}_L)] + b_v
   \]
\end{enumerate}

\subsection*{Detailed Training Algorithm}

\begin{enumerate}
\item \textbf{Initialization:}
   \begin{itemize}
   \item Initialize actor networks $\pi_{\phi_i}$ for each agent $i$
   \item Initialize centralized critic network $V_\theta$
   \item Set up experience buffer $\mathcal{D}$ with capacity $N$
   \item Initialize environment runners for parallel data collection
   \end{itemize}

\item \textbf{Main Training Loop:}
   \begin{enumerate}
   \item \textbf{Data Collection Phase:}
      \begin{itemize}
      \item Run $M$ parallel environments for $T$ timesteps
      \item For each timestep $t$ and agent $i$:
         \begin{itemize}
         \item Process observation $o_t^i$ through GRU and attention layers
         \item Sample action $a_t^i \sim \pi_{\phi_i}(\cdot|h_t^i)$
         \item Store tuple $(o_t^i, a_t^i, r_t^i, h_t^i)$ in buffer $\mathcal{D}$
         \end{itemize}
      \end{itemize}

   \item \textbf{Advantage Estimation:}
      \begin{itemize}
      \item Compute returns using GAE-$\lambda$:
      \[
      \hat{R}_t = r_t + \gamma\lambda r_{t+1} + (\gamma\lambda)^2r_{t+2} + \cdots
      \]
      \item Calculate advantages:
      \[
      \hat{A}_t = \hat{R}_t - V_\theta(s_t)
      \]
      \item Normalize advantages:
      \[
      \hat{A}_t \leftarrow \frac{\hat{A}_t - \mu_{\hat{A}}}{\sigma_{\hat{A}} + \epsilon}
      \]
      \end{itemize}

   \item \textbf{Policy Update Phase:}
      \begin{itemize}
      \item For $K$ epochs:
         \begin{itemize}
         \item Sample mini-batches of size $B$ from $\mathcal{D}$
         \item Compute policy loss:
         \[
         L_{\text{CLIP}}(\phi_i) = \mathbb{E}_t[\min(r_t(\phi_i)\hat{A}_t, \text{clip}(r_t(\phi_i), 1-\epsilon, 1+\epsilon)\hat{A}_t)]
         \]
         \item Compute value function loss:
         \[
         L_{\text{VF}}(\theta) = \frac{1}{2}\mathbb{E}_t[(V_\theta(s_t) - \hat{R}_t)^2]
         \]
         \item Compute attention losses:
         \begin{align*}
         L_{\text{self}} &= -\sum_t\sum_{k=1}^K \alpha_t^i(k)\log(\alpha_t^i(k)) \\
         L_{\text{cross}} &= -\sum_t\sum_{j\neq i} \beta_t^i(j)\log(\beta_t^i(j))
         \end{align*}
         \item Update parameters:
         \begin{align*}
         \phi_i &\leftarrow \phi_i + \eta_\pi\nabla_{\phi_i}(L_{\text{CLIP}} + c_1L_{\text{self}} + c_2L_{\text{cross}}) \\
         \theta &\leftarrow \theta - \eta_v\nabla_\theta L_{\text{VF}}
         \end{align*}
         \end{itemize}
      \end{itemize}
   \end{enumerate}
\end{enumerate}

\subsection*{Detailed Hyperparameter Settings}

\begin{enumerate}
\item \textbf{Network Architecture:}
   \begin{itemize}
   \item Input embedding dimension: 128
   \item GRU hidden dimension: 256
   \item Attention key/value dimension: 64
   \item Number of attention heads: 8
   \item Feed-forward hidden dimension: 512
   \item Layer normalization epsilon: $1 \times 10^{-5}$
   \end{itemize}

\item \textbf{Training Parameters:}
   \begin{itemize}
   \item Learning rate (actor): $3 \times 10^{-4}$
   \item Learning rate (critic): $1 \times 10^{-3}$
   \item Batch size: 64
   \item Number of epochs: 10
   \item Clipping parameter $\epsilon$: 0.2
   \item GAE parameter $\lambda$: 0.95
   \item Discount factor $\gamma$: 0.99
   \item Gradient clipping norm: 0.5
   \end{itemize}

\item \textbf{Attention-Specific:}
   \begin{itemize}
   \item History window size: 50
   \item Attention dropout rate: 0.1
   \item Temperature for attention softmax: 1.0
   \item Self-attention weight $c_1$: 0.1
   \item Cross-attention weight $c_2$: 0.1
   \end{itemize}

\item \textbf{Buffer Parameters:}
   \begin{itemize}
   \item Buffer capacity: $1 \times 10^6$
   \item Minimum buffer size for training: $1 \times 10^4$
   \item Number of parallel environments: 16
   \item Steps per environment: 128
   \end{itemize}
\end{enumerate}

\subsection*{Training Optimizations}

\begin{enumerate}
\item \textbf{Memory Management:}
   \begin{itemize}
   \item Use half-precision (FP16) training
   \item Implement gradient checkpointing
   \item Employ efficient attention computation using sparse attention
   \item Implement circular buffer for history management
   \end{itemize}

\item \textbf{Computational Optimizations:}
   \begin{itemize}
   \item Vectorize environment interactions
   \item Use GPU-accelerated tensor operations
   \item Implement parallel advantage computation
   \item Use efficient mini-batch sampling
   \end{itemize}

\item \textbf{Stability Techniques:}
   \begin{itemize}
   \item Gradient norm clipping
   \item Learning rate scheduling
   \item Value function clipping
   \item Entropy regularization
   \item Early stopping based on KL divergence
   \end{itemize}

\item \textbf{Curriculum Learning:}
   \begin{itemize}
   \item Progressive attention head annealing
   \item Gradual increase in history window size
   \item Adaptive batch size scheduling
   \item Dynamic learning rate adjustment
   \end{itemize}
\end{enumerate}
